---
title: "Measles Susceptibility in Glasgow Primary School Children"
author: "DAS Group 15"
date: today
format:
  pdf:
    toc: false
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
# load packages
library(tidyverse)
library(ggplot2)
library(broom)
library(knitr)
library(kableExtra)
library(sjPlot)
library(performance)
```

```{r}
dat <- read.csv('DAProject15.csv')

# create proportion column (Y susceptible out of N total)
dat$prop <- dat$Y / dat$N

# centre year at 2004 - makes the intercept easier to interpret
# (i.e. intercept = log-odds at 2004, not at year 0 which makes no sense)
dat$yr_c <- dat$Year - 2004

# indicator for post-retraction period
dat$post <- ifelse(dat$Year >= 2004, 1, 0)
```

# Introduction {#sec-intro}

Since the late 1980s, the measles, mumps, and rubella (MMR) vaccine has been the foundation of childhood vaccination in the UK. However, a now-discredited report by Andrew Wakefield in 1998 that proposed a connection between the vaccine and autism badly damaged public trust. Due to this false information, vaccination rates dropped to almost 80% in some areas by 2003, which subsequently caused serious measles epidemics.

Even if the Wakefield report was partially and fully retracted in 2004 and 2010, respectively, the timeframe for the restoration of public confidence and the ensuing decline in measles susceptibility is still crucial for analysis. This study assesses the measles susceptibility of primary school students in 133 intermediate zones (IZs) in Glasgow between 1998 and 2012 using data from the Scottish Childhood Immunisation Record System (SCIRS).

This study evaluates the probabilities of susceptibility in each post-1998 year directly against the 1998 baseline, avoiding the assumption of a linear trend by treating each year as a categorical component. In particular, we want to address:

1.Did Glasgow exhibit a statistically significant change in measles susceptibility relative to the 1998 baseline?

2.Did the turning point in susceptibility coincide with the 2004 partial retraction, or was there a measurable lag in public health recovery

<!--
Are we referencing for this report? If so, we will require some for this section.
Some interesting articles include:
- https://pmc.ncbi.nlm.nih.gov/articles/PMC3136032/ (National Institutes of Health)
- https://www.bmj.com/content/342/bmj.c7452 (British Medical Journal)
- https://www.chop.edu/vaccine-education-center/vaccine-safety/vaccines-and-other-conditions/autism (Children's Hospital of Philadelphia)
--> 

# Exploratory Data Analysis {#sec-exp}

<!--
Is the IZ column necessary? 
Would this table be more readable as a % rather than raw proportion?
-->

```{r}
# summarise by year - want mean proportion, sd, min, max per year
year_sum <- dat %>%
  group_by(Year) %>%
  summarise(
    n_iz     = n(),
    mean_p   = round(mean(prop), 4),
    sd_p     = round(sd(prop), 4),
    min_p    = round(min(prop), 4),
    max_p    = round(max(prop), 4)
  )
```

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for the proportion of primary school children susceptible to measles, by year."

kable(year_sum,
      col.names = c("Year", "IZs", "Mean", "SD", "Min", "Max"),
      booktabs  = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  add_header_above(c(" " = 2, "Proportion susceptible" = 4))
```

```{r}
#| label: tbl-summary2
#| tbl-cap: "Summary statistics for the proportion of primary school children susceptible to measles, by year."

year_sum_percentages <- cbind(year_sum[,1], year_sum[,c(2:6)]*100)
kable(year_sum_percentages[, -c(2)],
      col.names = c("Year", "Mean", "SD", "Min", "Max"),
      booktabs  = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  add_header_above(c(" " = 1, "Percenatge susceptible (%)" = 4))
  
```

@tbl-summary shows how the mean proportion of susceptible children
changed over time. There is an increase from 1998 through to around
2003, followed by a decline from 2004 onwards, which lines up with the
timing of the Wakefield retraction.

```{r}
#| label: fig-trend
#| fig-cap: "Mean proportion susceptible per year across all 133 Glasgow IZs. Error bars show +/- 1 standard error. The dashed line marks 2004."
#| fig-width: 6
#| fig-height: 3.2
#| fig-align: center

ggplot(year_sum, aes(x = Year, y = mean_p)) +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue", size = 2) +
  geom_errorbar(aes(ymin = mean_p - sd_p/sqrt(n_iz),
                    ymax = mean_p + sd_p/sqrt(n_iz)),
                width = 0.3, colour = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 2004, linetype = "dashed", colour = "firebrick") +
  annotate("text", x = 2004.2, y = max(year_sum$mean_p) * 0.99,
           label = "2004 retraction", hjust = 0, size = 3, colour = "firebrick") +
  labs(x = "Year", y = "Mean proportion susceptible") +
  theme_minimal()
```

<!--
Our explanatory analysis only addresses question 1. We could also 
talk about how the downward trend only starts to take place after 2008, not 
specifically 2004 as asked in question 2. Seems to imply it took at least 4 years
for an uptake in MMR vaccines post partial retraction.
-->

@fig-trend shows the overall pattern of susceptibility rising from 1998,
peaks around 2003, then falls after 2004. This is consistent with the
drop in vaccination rates following the 1998 paper, and a recovery in
uptake after the retraction. However, the change in the overall trend does not 
occur in 2004 directly. There appears to be an overall increase in susceptibility
until 2008. That is, there appears to be a delayed effect.

<!--
Vertical line for 2004 appears to be misaligned when rendered. Moving it to 2004 
blocks some of the information. Maybe remove it for this plot?
-->

```{r}
#| label: fig-box
#| fig-cap: "IZ-level proportion susceptible by year. Each box summarises the 133 IZs for that year."
#| fig-width: 7
#| fig-height: 3.2
#| fig-align: center

ggplot(dat, aes(x = factor(Year), y = prop)) +
  geom_boxplot(fill = "lightsteelblue", outlier.size = 0.7, outlier.alpha = 0.4) +
#  geom_vline(xintercept = 6.5,   # 2004 is the 7th year (1998,99,00,01,02,03,2004)
 #            linetype = "dashed", colour = "firebrick") +
  labs(x = "Year", y = "Proportion susceptible") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@fig-box shows the distribution across IZs for each year. There is
moderate variation between IZs, which makes sense given that different
areas will have different vaccination uptake.

# Formal Analysis {#sec-formal}

## Model

Since our outcome is a count of susceptible children out of a total
(i.e., Y out of N), the model proposed is a binomial GLM with a logit
link, which is a logistic regression model but for proportional count
data rather than individual 0/1 outcomes.

We want to test whether the trend in susceptibility changed at 2004, so
we fit a model with an interaction term between year and the post-2004
indicator:

<!--
We should probably directly introduce the pi variable as the our proportion (y_i/n_i).
-->

$$
\log\left(\frac{\pi_i}{1-\pi_i}\right) = \beta_0 + \beta_1 \cdot \text{Year}_i^* + \beta_2 \cdot \mathbb{1}[\text{post}] + \beta_3 \cdot (\text{Year}_i^* \times \mathbb{1}[\text{post}])
$$
$$
\log\left(\frac{\pi_i}{1-\pi_i}\right) = \beta_0  + \beta_1 \cdot \mathbf{1}[\text{2000}] + 
\beta_2 \cdot \mathbf{1}[\text{2002}] + \beta_3 \cdot \mathbf{1}[\text{2004}] + \beta_4 \cdot \mathbf{1}[\text{2008}]
+ \beta_5 \cdot \mathbf{1}[\text{2010}] + \beta_6 \cdot \mathbf{1}[\text{2012}]
$$

where $\text{Year}^*$ is the year centred at 2004, and
$\mathbb{1}[\text{post}]$ equals 1 for 2004 and later. The interaction
$\beta_3$ captures whether the slope differs before vs after 2004. Here, 
$\pi_i$ is the probability of a child being susceptible to measles in year level $i$.

Fitting the this proposed model produced the following summary output. 

```{r}
#| echo: false
#| tbl-cap: "Table 2: Model summary for binomial model centred at 2004."

model_binom <- glm(cbind(Y,N-Y) ~ yr_c + post + yr_c*post, family = binomial, dat)# Fit Binomial model

# Model summary
tab_model(model_binom, show.aic = T, show.ci = F, show.dev = T)

```
<!--
Checking for overdispersion. 
-->
```{r}
#| echo: false

# Overdispersion diagnostic 
# Ratio > 1 suggests overdispersion, meaning binomial SEs
# may be too narrow and p-values too optimistic
disp_ratio <- deviance(model_binom) / df.residual(model_binom)
cat("Dispersion ratio:", round(disp_ratio, 3), "\n")

check_overdispersion(model_binom)#Alternatively

```

The residual deviance of the binomial model (2044.5 on 927 df) gives a dispersion ratio of approximately 2.2, which exceeds the expected value of 1 under a correctly specified binomial model. This suggests the presence of overdispersion, meaning the standard errors from the binomial model may be too narrow. 


<!--
TB
difference in deviance (null - residual deviance), using a chi-squared dis to test if significant.
-->

```{r}
#| echo: false
#| output: false
pchisq((2179.6-2044.5), df = 3, lower.tail = FALSE)
```
<!--
TB
possible answer of code above: Model fit improved substantially compared with the null model, with deviance decreasing from 2179.6 (930 df) to 2044.5 (927 df). This reduction of $\chi^2 = 135.1$ with 3 degrees of freedom is highly significant ($p < 0.05$), indicating the predictors significantly improve the model fit.
-->

<!--
TB
Summary gives us exact coefficients
-->
```{r}
#| echo: false
#| output: false
summary(model_binom)
```

<!--
Need to interpret these coefficients.
-->

<!--
(TB - possible answer to the coefficients interpretation. Word count could be cut??) 
The binomial logistic regression model suggests that measles susceptibility changed significantly over time and around the 2004 intervention point. Before 2004, the positive coefficient for `yr_c` ($\beta = 0.112$, $p < 0.001$) indicates that the log-odds of susceptibility were increasing each year. The coefficient for `post` ($\beta = -0.206$, $p = 0.001$) suggests a significant immediate decrease in susceptibility after 2004. Furthermore, the interaction term `yr_c:post` ($\beta = -0.162$, $p < 0.001$) indicates that the time trend changed after 2004, with the slope becoming negative. Overall, this suggests that susceptibility was rising before 2004 but began to decline slightly afterwards, indicating a significant change in trend following the retraction of the Wakefield article.
-->


<!--
An idea for addressing question 2 is by moving our post covariate to mark years after 
2008 rather than 2004. That is, there seems to be a greater decline in susceptibility 
after 2008. By doing some model comparisons, we can infer that the change in measles 
susceptibility occurred post 2008.
I've included what such an alternative model may look like.
-->

<!--
TB
Included Deviance
discussed deviance with repsect to previous model. This is below the code
-->

```{r}
#| echo: false
#| tbl-cap: "Table 3: Model summary for binomial model centred at 2008."

# Define alternative dataset centred around 2008.
dat_alt <- dat
dat_alt$post <- ifelse(dat_alt$Year >= 2008, 1, 0)
dat_alt$yr_c <- dat_alt$yr_c - 4

# Fit model
model_binom_alt <- glm(cbind(Y,N-Y) ~ yr_c + post + yr_c*post, family = binomial, data = dat_alt)

# Model summary
tab_model(model_binom_alt, show.ci = F, show.aic = T, show.dev = T)
```

<!--
TB (possible discussion around comparing deviances and AIC for the model_binom and model_binom_alt)
The model centred at 2004 has a residual deviance of 2044.5 and AIC of 5156.5, while the model centred at 2008 has a lower residual deviance of 2002.9 and AIC of 5115. Since both the deviance and AIC are smaller for the model centred at 2008, it provides a better fit to the data. This suggests that the change in measles susceptibility alogns more with a shift occurring closer to 2008 rather than exactly in 2004 (reinforcing the possible lag in sentiment between the article retraction and actual increase in vacinations). However, both models show statistically significant effects for the post-intervention term and the change in slope.
-->








<!--
TB (couldn't find it, so useless on github, hope i didn't mess up anything)
Model with year as factor and analysis below
-->
## Modelling Measles Susceptibility by Year

### Binomial Logistic Regression Model

To assess variation in measles susceptibility across years, a binomial logistic regression model was fitted:

$$
Y_i \sim \text{Binomial}(N_i, p_i)
$$

$$
\log\left(\frac{p_i}{1 - p_i}\right) = \beta_0 + \beta_1 \cdot \text{Year}_{2000,i} + \beta_2 \cdot \text{Year}_{2002,i} + \cdots + \beta_6 \cdot \text{Year}_{2012,i}
$$

where \(p_i\) is the probability that a child in area \(i\) is susceptible to measles, and 1998 is treated as the reference category.

```{r}
#| include: false
#| tbl-cap: "Table 4: Odds Ratios for Year-on-Year Susceptibility Relative to 1998 Baseline"
dat$Year <- as.factor(dat$Year)
model_year <- glm(
  cbind(Y, N - Y) ~ Year,
  family = binomial,
  data = dat
)

tab_model(model_year, show.ci = F, show.aic = T, show.dev = T)
```

```{r}
#| echo: false
#| eval: false
summary(model_year)
```

```{r}
#| include: false
dispersion <- model_year$deviance / model_year$df.residual
dispersion
```

$$\hat{\phi} = \frac{\text{Residual Deviance}}{\text{Residual Degrees of Freedom}} = `r round(dispersion, 3)`$$
Overdispersion was assessed using the binomial model, with the ratio of residual deviance to degrees of freedom exceeding 1, indicating the presence of over dispersion. To account for this, a quasibinomial model was fitted, which adjusts the standard errors to provide more reliable inference.

To account for overdispersion, a quasibinomial model was fitted:
$$
Y_i \sim \text{Quasi-Binomial}(N_i, p_i, \phi)
$$

$$
\log\left(\frac{p_i}{1 - p_i}\right)=\beta_0 + \beta_1 \cdot \text{Year}_{2000,i} + \cdots + \beta_6 \cdot \text{Year}_{2012,i}
$$

```{r}
#| echo: false
#| tbl-cap: "Odds Ratios for Year-on-Year Susceptibility Relative to 1998 Baseline"
model2 <- glm(cbind(Y, N - Y) ~ Year, 
                   data = dat, 
                   family = quasibinomial)

tab_model(model_year, show.ci = F, show.aic = T, show.dev = T)
```

The year-specific logistic regression model shows that measles susceptibility varied significantly across years in Glasgow. Compared with the baseline year (1998), susceptibility increased significantly in 2000, 2002, 2004 and 2008, with the largest increase occurring in 2008 (OR = 1.70, $p < 0.001$). However, there was no significant difference in 2010 or 2012. This suggests that Glasgow did experience changes in measles susceptibility over time.

However, the results do not indicate a clear change occurring specifically in 2004, the year of the Wakefield article retraction. Instead, susceptibility appears to continue increasing after 2004 and peaks around 2008, suggesting the effect was not immediate.

```{r}
#| include: false
summary(model2)
```

```{r}
#| include: false
pchisq(2179.6 - 1966.6, df = 930 - 924, lower.tail = FALSE)
```

To determine if "Year" is a significant predictor of susceptibility, we compared our model against a null model using a likelihood ratio test:

$$
\chi^2 = D_{\text{null}} - D_{\text{model}} = 2179.6 - 1966.6 = 213.0
$$

$$
df = 930 - 924 = 6
$$

$$
p = P\left(\chi^2_6 \ge 213.0\right) \approx 3.23 \times 10^{-43}
$$
The reduction in deviance from 2179.6 to 1966.6 corresponds to $\chi^2 = 213.0$ with 6 degrees of freedom ($p \approx 3.23 \times 10^{-43}$), providing strong evidence that including year significantly improves model fit, confirming that Glasgow exhibited a clear change in measles susceptibility between 1998 and 2012. This result is identical for both the binomial and quasibinomial models, as the quasibinomial approach adjusts only the standard errors and not the fitted values or deviance.


```{r}
#| echo: false
#| eval: false

print(check_model(model_year))
print(check_model(model_binom))

check_model(model_binom,
            check = c("pp_check", "binned_residuals", "vif"))

check_model(model_year,
            check = c("pp_check", "binned_residuals", "qq", "influential_observations"))

```

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
print(check_model(model_year))
```
The diagnostic plots indicate that the year-based binomial model provides an adequate fit to the data. The posterior predictive check shows good agreement between observed and fitted distributions, while binned residuals exhibit no systematic deviations. There is no evidence of influential observations, and residuals appear well-calibrated. Minor discrepancies in the upper tail suggest slight underestimation of extreme values, but these do not materially affect the overall model fit. These plots are the same for the year-based quasibinomial model.


# Consclusion {#sec-conc}
<!--
TB(brief conclusion)
The results indicate that measles susceptibility varied significantly across years. Compared with 1998, susceptibility increased significantly in the early 2000s and peaked around 2008, before declining in later years. Evidence of overdispersion suggests that the quasibinomial model provides more reliable standard errors, but the overall conclusions remain unchanged
-->






